Towards a Verified Tableau Prover for a Ground Fragment of Set Theory

### 2022-11-23 paper/paper.tex
Work in comments from reviewers

%%%%%%%%%\listoftodos{}

\section{Introduction}
In contrast to pen-and-paper proofs where we expect the reader to fill in trivial details, interactive theorem provers (ITPs) require us to justify each proof step rigorously.
Manually reducing each proof step down to the axioms would be impractical, though;
hence, ITPs usually offer a wide array of automated methods to deal with the proof goals that arise during interactive use.

@@

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitation is provided by Lawrence in an entry~\cite{zfc_in_hol_afp} of the \textit{Archive of Formal Proofs}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
Two articles~\cite{zfc_in_hol1,zfc_in_hol2} have been written that provide further context to this entry.
The entry declares a type \lstinline!V! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!elts :: V $\Rightarrow$ V set! maps a set to its elements.
      Note that even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class (relative to \lstinline!V!).
      This class only constitutes a set (relative to \lstinline!V!) if the class is small.
  \item The predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, which is polymorphic in \lstinline!'a!, indicates whether a class is small.
  \item The function \lstinline!vset :: 'a set $\Rightarrow$ V! turns a class into a set of type \lstinline!V! if it is small. 
\item The usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: V!.
\end{itemize}

@@

    \lstinline!@\fmNeg{(\fmOr{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! &&
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\

    \lstinline!@\fmNeg{(\fmOr{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! &&
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p! &&&& \\[3ex]

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation as displayed in \autoref{fig:alg} uses a couple features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises for which arguments the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
To be able to reason about this predicate, we pass the \lstinline!domintros! flag that instructs the function package to derive introduction rules for it.
In \autoref{sec:correct}, we will use these rules to discharge the guarding assumptions for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition.

@@

\section{Completeness of the Calculus\label{sec:complete}}
For completeness of the calculus, we need to show that every unsatisfiable formula has a closed tableau or, conversely, that the formula is satisfiable if there is a saturated and open branch in the tableau.
To facilitate inductive reasoning, we show a stronger statement by constructing a model \lstinline!M! such that \lstinline!M $\models$ $\phi$! for all \lstinline!$\phi$ $\in$ set b!.
At the core of the model, there is a \textit{realisation} function that maps set terms to sets of type \lstinline!V!.
A subset of the witnesses, which we call \textit{pure} witnesses, receives special treatment from the realisation function for reasons that will become apparent in \autoref{sec:pwits}.
The set terms of a branch can thus be partitioned into two collections as defined below.
\begin{lstlisting}
definition wits' :: 'a branch $\Rightarrow$ 'a set where
  wits' b $\equiv$
    {c $\in$ wits b. $\forall$t $\in$ subterms (last b).
      AT (Var c $\eqS$ t) $\notin$ set b $\land$
      AT (t $\eqS$ Var c) $\notin$ set b}

definition subterms' ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms' b $\equiv$
    subterms (last b) $\cup$ 
    Var ` (wits b - wits' b)
\end{lstlisting}

We aim to construct a syntactic model that is derived from the membership literals \lstinline!s $\inS$ t! in the branch.
To this end, we construct a graph whose vertices are the disjoint union of the sets above and there is an edge from \lstinline!s! to \lstinline!t! in the graph if, and only if, \lstinline!s $\inS$ t! is in \lstinline!b!.
Note that we use \citeauthor{graph_theory_afp}'s graph library~\cite{graph_theory_afp} that represents a graph as a record of vertices, arcs (directed edges), and two functions \lstinline!tail! and \lstinline!head! that map an arc to its source respectively target vertex.
\begin{lstlisting}
definition bgraph b $\equiv$
  let vs = Var ` wits' b $\cup$ subterms' b
  in $\llparenthesis$ verts = vs,
     &$\phantom{\llparenthesis}$& arcs = {(s, t). (s $\inS$ t) $\in$ set b},
     &$\phantom{\llparenthesis}$& tail = fst, head = snd $\rrparenthesis$
\end{lstlisting}

The realisation function is defined relative to this graph.
As mentioned before, the pure witnesses \lstinline!Var ` wits' b! and the rest of the set terms \lstinline!subterms' b! are treated differently by the realisation function.
Terms in the latter set are evaluated in accordance to the structure of the graph, i.e.\ the realisation of a vertex is defined as the union of the realisations of the parent vertices. 
For the former set, we choose a function \lstinline!I! that assigns the witnesses in \lstinline!Var ` wits' b! pairwise distinct sets with cardinality greater than that of the vertices.
We can always choose such a function since we assume an infinite universe of variables.
Then, we return the singleton set \lstinline!vset {I x}!, which, together with the cardinality constraint, guarantees that realisations are distinct between pure witnesses themselves as well as between pure witnesses and set terms.  
The notation \lstinline!u $\rightarrow_\text{G}$ s! in the definition below indicates that there is an edge from \lstinline!u! to \lstinline!s! in the graph \lstinline!G!.
\begin{lstlisting}
abbreviation parents G s $\equiv$ {u. u $\rightarrow_\text{G}$ s}

function realise :: 'a pset_term $\Rightarrow$ V where
  x $\in$ Var ` wits' b $\Longrightarrow$ realise x = vset {I x}
| x $\in$ subterms' b
  $\Longrightarrow$ realise t =
  &\phantom{$\Longrightarrow$ }&  vset {realise ` parents (bgraph b) s}
| x $\notin$ verts G $\Longrightarrow$ realise x = 0
\end{lstlisting}

@@

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses of a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realisation of a pure witness does not depend on the realisation of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b
  assumes c $\in$ wits' b
  shows (Var c $\eqS$ t) $\notin$ set b
    and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realise!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: V!, were they not treated separately. 
To see that this is a problem, consider the branch
\begin{lstlisting}
  b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]
\end{lstlisting}
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y,
   Var x $\inS$ Var s, Var x $\notinS$ Var t,
   Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.

@@

\subsection{Realisation of an Open Branch\label{sec:real}}
Remember that for completeness, we need to show that the realisation function for an open and saturated branch \lstinline!b! actually constitutes a model for all formulae in the branch.
We start by verifying that the realisation function models all literals in the branch; more formally, the following propositions hold:
\begin{enumerate}[label={(\roman*)}]
  \item\label{it:at_mem} We have \lstinline!realise s $\in$ elts (realise t)! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realise s $=$ realise t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realise s $\neq$ realise t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realise s $\notin$ elts (realise t)! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \hyperref[lst:lemma_2]{\lstinline!lemma_2!}, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle, so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ wits' b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \hyperref[lst:lemma_2]{\lstinline!lemma_2!}.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\inS$ subterms' b! must hold.
  Now, assume for contradiction that
\begin{lstlisting}
  realise s $\neq$ realise t
\end{lstlisting}
  which implies that
\begin{lstlisting}
  elts (realise s) $\neq$ elts (realise t).
\end{lstlisting}
  Without loss of generality --- the other case is symmetric --- we obtain an \lstinline!e! such that
\begin{lstlisting}
  e $\in$ elts (realise s) $\land$ e $\notin$ elts (realise t). 
\end{lstlisting}
  Considering the fact that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realise!, we obtain an \lstinline!u! where \lstinline!e = realise u! and \lstinline!u $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!u $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ set b! and the saturation of \lstinline!b!, it follows that \lstinline!u $\inS$ t! must also be in \lstinline!b!.
  But then we have
\begin{lstlisting}[columns=fixed]
  realise u $\in$ elts (realise t)
        = e $\in$ elts (realise t)
\end{lstlisting}
  using Proposition~\ref{it:at_mem}, which stands in contradiction to the assumption \lstinline!e $\notin$ elts (realise t)!.
\end{proof}

@@

\noindent To finish of the proof of total correctness, note that every singleton branch is trivially well-formed; thus, termination, completeness, and soundness easily transfer to \lstinline!mlss_proc!.
\begin{lstlisting}
theorem mlss_proc_complete:
  fixes $\phi$ :: 'a pset_fm
  assumes $\neg$ mlss_proc $\phi$
  assumes infinite (UNIV :: 'a set)
  shows $\exists$M. M $\models$ $\phi$

>>>>

\listoftodos{}

\section{Introduction}
In contrast to pen-and-paper proofs where we expect the reader to fill in trivial details, interactive theorem provers (ITPs) require us to justify each proof step rigorously.
\todo[inline]{Reviewer: Reformulate below because we can use cuts}
Manually reducing each proof step down to the axioms would be impractical, though;
hence, ITPs usually offer a wide array of automated methods to deal with the proof goals that arise during interactive use.

@@

In Isabelle/HOL, which is simply typed, this definition is not accepted, though.
A way to work around this limitation is provided by \citeauthor{zfc_in_hol_afp} in an entry~\cite{zfc_in_hol_afp} of the \textit{Archive of Formal Proofs}, which adds $\mathcal{V}$ to Isabelle/HOL by way of axiomatisation.
Two articles~\cite{zfc_in_hol1,zfc_in_hol2} have been written that provide further context to this entry.
The entry declares a type \lstinline!V! that comes with the following functionality:
\begin{itemize}
  \item The function \lstinline!elts :: V $\Rightarrow$ V set! maps a set to its elements.
      Note that even though we use the typed sets provided by Isabelle/HOL here, \lstinline!elts! actually returns a class (relative to \lstinline!V!).
      This class only constitutes a set (relative to \lstinline!V!) if the class is small.
  \item The predicate \lstinline!small :: 'a set $\Rightarrow$ bool!, which is polymorphic in \lstinline!'a!, indicates whether a class is small.
  \item The function \lstinline!vset :: 'a set $\Rightarrow$ V! turns a class into a set of type \lstinline!V! if it is small. 
\item The usual set operations such as equality ($=$), union ($\sqcup$), intersection ($\sqcap$), and difference ($-$) are defined.
\item Finally, the empty set coincides with the ordinal $0$, so it is denoted by \lstinline!0 :: V!.
\end{itemize}

@@

    \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, p! & $\Longrightarrow$ & \lstinline!@\fmNeg{q}@! &&
    \lstinline!s $\inS$ $\ton$ $\unionS$ $\ttw$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\inS$ $\ton$! \\

    \lstinline!@\fmNeg{(\fmAnd{p}{q})}@, q! & $\Longrightarrow$ & \lstinline!@\fmNeg{p}@! &&
    \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! \\
    \lstinline!@\fmNeg{(\fmNeg{p})}@! & $\Longrightarrow$ & \lstinline!p! &&&& \\[3ex]

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation as displayed in \autoref{fig:alg} uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises for which arguments the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
To be able to reason about this predicate, we pass the \lstinline!domintros! flag that instructs the function package to derive introduction rules for it.
In \autoref{sec:correct}, we will use these rules to discharge the guarding assumptions for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition.

@@

\section{Completeness of the Calculus\label{sec:complete}}
For completeness of the calculus, we need to show that every unsatisfiable formula has a closed tableau or, conversely, that the formula is satisfiable if there is a saturated and open branch in the tableau.
To facilitate inductive reasoning, we show a stronger statement by constructing a model \lstinline!M! such that \lstinline!M $\models$ $\phi$! for all \lstinline!$\phi$ $\in$ set b!.
At the core of the model, there is a \textit{realisation} function that maps set terms to sets of type \lstinline!V!.
A subset of the witnesses, which we call \textit{pure} witnesses, receives special treatment from the realisation function for reasons that will become apparent in \autoref{sec:pwits}.
The collection of set terms of a branch can thus be partitioned into two collections as defined below.
\begin{lstlisting}
definition pwits :: 'a branch $\Rightarrow$ 'a set where
  pwits b $\equiv$
    {c $\in$ wits b. $\forall$t $\in$ subterms (last b).
      AT (Var c $\eqS$ t) $\notin$ set b $\land$
      AT (t $\eqS$ Var c) $\notin$ set b}

definition subterms' ::
    'a branch $\Rightarrow$ 'a pset_term set where
  subterms' b $\equiv$
    subterms (last b) $\cup$ 
    Var ` (wits b - pwits b)
\end{lstlisting}

We aim to construct a syntactic model that is derived from the membership literals \lstinline!s $\inS$ t! in the branch.
To this end, we construct a graph whose vertices are the disjoint union of the sets above and there is an edge from \lstinline!s! to \lstinline!t! in the graph if, and only if, \lstinline!s $\inS$ t! is in \lstinline!b!.
Note that we use \citeauthor{graph_theory_afp}'s graph library~\cite{graph_theory_afp} that represents a graph as a record of vertices, arcs (directed edges), and two functions \lstinline!tail! and \lstinline!head! that map an arc to its source, respectively target, vertex.
\begin{lstlisting}
definition bgraph b $\equiv$
  let vs = Var ` pwits b $\cup$ subterms' b
  in $\llparenthesis$ verts = vs,
     &$\phantom{\llparenthesis}$& arcs = {(s, t). (s $\inS$ t) $\in$ set b},
     &$\phantom{\llparenthesis}$& tail = fst, head = snd $\rrparenthesis$
\end{lstlisting}

The realisation function is defined relative to this graph.
As mentioned before, the pure witnesses \lstinline!Var ` pwits b! and the rest of the set terms \lstinline!subterms' b! are treated differently by the realisation function.
Terms in the latter set are evaluated in accordance to the structure of the graph, i.e.\ the realisation of a vertex is defined as the union of the realisations of the parent vertices. 
For the former set, we choose a function \lstinline!I! that assigns the witnesses in \lstinline!Var ` pwits b! pairwise distinct sets with cardinality greater than that of the vertices.
We can always choose such a function since we assume an infinite universe of variables.
Then, we return the singleton set \lstinline!vset {I x}!, which, together with the cardinality constraint, guarantees that realisations are distinct between pure witnesses themselves as well as between pure witnesses and set terms.  
The notation \lstinline!u $\rightarrow_\text{G}$ s! in the definition below indicates that there is an edge from \lstinline!u! to \lstinline!s! in the graph \lstinline!G!.
\begin{lstlisting}
abbreviation parents G s $\equiv$ {u. u $\rightarrow_\text{G}$ s}

function realise :: 'a pset_term $\Rightarrow$ V where
  x $\in$ Var ` pwits b $\Longrightarrow$ realise x = vset {I x}
| x $\in$ subterms' b
  $\Longrightarrow$ realise t =
  &\phantom{$\Longrightarrow$ }&  vset {realise ` parents (bgraph b) s}
| x $\notin$ verts G $\Longrightarrow$ realise x = 0
\end{lstlisting}

@@

\subsection{Characterisation of the Pure Witnesses\label{sec:pwits}}
Recall that the pure witnesses of a branch \lstinline!b! are those witnesses that are not related to other subterms in \lstinline!last b! by equality.
In the context of a well-formed branch, this characterisation can be strengthened to any set term and, in addition, we also get that there is no membership literal where a pure witness is on the right-hand side.
Intuitively speaking, the realisation of a pure witness does not depend on the realisation of any other set term.
\begin{lstlisting}[label={lst:lemma_2}]
lemma lemma_2:
  assumes wf_branch b
  assumes c $\in$ pwits b
  shows (Var c $\eqS$ t) $\notin$ set b
    and (t $\eqS$ Var c) $\notin$ set b
    and (t $\inS$ Var c) $\notin$ set b
\end{lstlisting}
So why are pure witnesses treated differently?
According to the definition of \lstinline!realise!, the pure witnesses would be evaluated to the empty set \lstinline!0 :: V!, were they not treated separately. 
To see that this is a problem, consider the branch
\begin{lstlisting}
  b = [Var s $\neqS$ Var t, Var t $\neqS$ Var u]
\end{lstlisting}
which expands to several open and saturated branches, one of which is
\begin{lstlisting}
  [Var x $\neqS$ Var y,
   Var x $\inS$ Var s, Var x $\notinS$ Var t,
   Var y $\inS$ Var t, Var y $\notinS$ Var u] @ b
\end{lstlisting}
for some fresh \lstinline!x! and \lstinline!y!.
Assigning both \lstinline!Var x! and \lstinline!Var y! a value of \lstinline!0! would contradict the literal \lstinline!Var x $\neqS$ Var y!.
To prevent this, we assign the pure witnesses pairwise different values.

@@

\subsection{Realisation of an Open Branch\label{sec:real}}
Remember that for completeness, we need to show that the realisation function for an open and saturated branch \lstinline!b! actually constitutes a model for all formulae in the branch.
We start by verifying that the realisation function models all literals in the branch; more formally, the following propositions hold:
\begin{enumerate}[label={(\roman*)}]
  \item\label{it:at_mem} We have \lstinline!realise s $\in$ elts (realise t)! if it holds that \lstinline!s $\inS$ t! is in \lstinline!b!. 
  \item\label{it:at_eq} We have \lstinline!realise s $=$ realise t! if \lstinline!s $\eqS$ t! is in \lstinline!b!.
  \item\label{it:af_eq} We have \lstinline!realise s $\neq$ realise t! if \lstinline!s $\neqS$ t! is in \lstinline!b!.
  \item\label{it:af_mem} We have \lstinline!realise s $\notin$ elts (realise t)! if it holds that \lstinline!s $\notinS$ t! is in \lstinline!b!. 
\end{enumerate}
To illustrate the usefulness of \hyperref[lst:lemma_2]{\lstinline!lemma_2!}, we prove Proposition~\ref{it:at_eq}.
The proofs of all propositions translate well into Isabelle, so we refer to the original paper~\cite{new_fast_tableau} for the remaining proofs. 
\begin{proof}[Proof of Proposition~\ref{it:at_eq}]
  Assume that \lstinline!s $\eqS$ t! is in \lstinline!b!.
  If there exists a \lstinline!c $\in$ pwits b! where \lstinline!s = Var c! or \lstinline!t = Var c!, we arrive at a contradiction due to \hyperref[lst:lemma_2]{\lstinline!lemma_2!}.
  Therefore, both \lstinline!s $\in$ subterms' b! and \lstinline!t $\in$ subterms' b! must hold.
  Now, assume for contradiction that
\begin{lstlisting}
  realise s $\neq$ realise t
\end{lstlisting}
  which implies that
\begin{lstlisting}
  elts (realise s) $\neq$ elts (realise t).
\end{lstlisting}
  Without loss of generality --- the other case is symmetric --- we obtain an \lstinline!e! such that
\begin{lstlisting}
  e $\in$ elts (realise s) $\land$ e $\notin$ elts (realise t). 
\end{lstlisting}
  Considering the fact that \lstinline!s $\in$ subterms' b! and the definition of \lstinline!realise!, we obtain an \lstinline!u! where \lstinline!e = realise u! and \lstinline!u $\rightarrow_\text{bgraph b}$ s!.
  This, in turn, yields that \lstinline!u $\inS$ s! must be in \lstinline!b!.
  Together with the assumption \lstinline!(s $\eqS$ t) $\in$ set b! and the saturation of \lstinline!b!, it follows that \lstinline!u $\inS$ t! must also be in \lstinline!b!.
  But then we have
\begin{lstlisting}[columns=fixed]
  realise u $\in$ elts (realise t)
        = e $\in$ elts (realise t)
\end{lstlisting}
  using Proposition~\ref{it:at_mem}, which stands in contradiction to the assumption \lstinline!e $\notin$ elts (realise t)!.
\end{proof}

@@

\noindent To finish off the proof of total correctness, note that every singleton branch is trivially well-formed; thus, termination, completeness, and soundness easily transfer to \lstinline!mlss_proc!.
\begin{lstlisting}
theorem mlss_proc_complete:
  fixes $\phi$ :: 'a pset_fm
  assumes $\neg$ mlss_proc $\phi$
  assumes infinite (UNIV :: 'a set)
  shows $\exists$M. M $\models$ $\phi$
